---
title: "CRISP-DM Skills Check (Penguins)"
author: "Add your name here"
output:
html_document:
toc: true
toc_depth: 2
number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(42)

#Run once and then comment out:
#install.packages(c("pandoc"))
#library(pandoc)
#pandoc_install("3.7.0.2")

#Run once and then comment out:
#install.packages(c("palmerpenguins","tidyverse","ggplot2"))
#library(palmerpenguins)
#library(tidyverse)
#library(ggplot2)

# Add any other libraries you need here:

```

# Executive summary (3–5 sentences)
*Please enter what you asked, what you did, key results, one implication here.*

# Framing the Problem
- **Question:** *Please enter your question here.*
- **Data:** Palmer Penguins dataset (3 species, 7 variables, 344 rows).
- **Modeling task:** *Please enter your task here.*
- **Success:** *Please enter the metric to measure the success of your model here.*

# Data Understanding
Load data and show its structure. Minimum requirements:

- Analysis of missing values.
- 2 visualizations with takeaways.

```{r}
# Load data
data <- data("penguins", package = "palmerpenguins")
str(penguins)
```

# Explorative Data Analysis
Minimum requirements:

- Distribution of numeric variables
- Relationship between 2 numeric variables
```{r}

```

**Takeaways (bullets):**

- Pattern 1: *Please enter your observations here.*
- Pattern 2: *Please enter your observations here.*
- Missing values/outliers: *Please enter your observations here.*

# Data Preparation
Minimum requirements:

- Drop or impute missing values (explain why)
- Convert categorical variables to factors
```{r}

```
*Note your choices in 1–2 sentences (what you dropped/kept and why).*

# Modeling (pick classification or clustering and delete the other section)

## Classification
Minimum requirements:

- Train/test split (80/20)
- One classification model (e.g., logistic regression, decision tree)
- Report KPI to evaluate
```{r}

```

*1–2 sentences: which classes confuse and why?*

## Clustering
Minimum requirements:

- Definition of number of clusters
- One clustering model (e.g., k-means, hierarchical)
- Report KPI to evaluate
```{r}

```
*1–2 sentences: profile each cluster.*

# Evaluation & Validation
Minimum requirements:

- **Metric:** *Please report final metric on test set.*
- **Interpretation (2–4 sentences):** *Please enter the strengths, failure modes (e.g., which species are confused), class imbalance etc here.*

# AI Assistance (brief)
- **Tools used:** *Please enter the tools you used (e.g., GitHub Copilot, ChatGPT) here.*
- **One prompt:** *Please paste one representative prompt that led to useful code here.*
- **How validated:** *Please describe how you checked correctness (e.g., sanity checks on metrics/plots) of the prompt outputs here.*