---
title: "CRISP-DM Skills Check (Penguins)"
author: "Alona, Karoline, Julio"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(42)

# selects the mirror.
chooseCRANmirror(ind=81)

#Run once and then comment out:
# install.packages(c("pandoc"))
# library(pandoc)
# pandoc_install("3.7.0.2")

#Run once and then comment out:
# install.packages(c("palmerpenguins","tidyverse","ggplot2", "caret", "rpart", "rpart.plot", "randomForest", "e1071", "class", "kknn", "gbm", "xgboost", "nnet", "pROC", "MLmetrics"))

# load libraries
library(palmerpenguins)
library(tidyverse)
library(ggplot2)

# TODO: use only libraries needed for this work and remove unused ones.
# Add any other libraries you need here:
library(caret)          # train/test split, confusionMatrix,unified model interface
library(rpart)          # Decision Tree
library(rpart.plot)     # nice tree visualization
library(randomForest)   # Random Forest
library(e1071)          # Naïve Bayes + SVM
library(class)          # KNN (basic version)
library(kknn)           # KNN via caret
library(gbm)            # Gradient Boosting Trees
library(xgboost)        # alternative gradient boosting (faster, more popular)
library(nnet)           # simple Neural Networks (built-in style)

# Evaluation extras
library(pROC)           # ROC curves (slide 11 of your lecture)
library(MLmetrics)      # F1 score helper if confusionMatrix isn't enough

```

# Executive summary (3–5 sentences)
*Please enter what you asked, what you did, key results, one implication here.*

# Framing the Problem
- **Question:** *To what extent can physical measurements predict the sex of Gentoo penguins, and which features are most informative?*
- **Data:** Palmer Penguins dataset (3 species, 7 variables, 344 rows).
- **Modeling task:** *Please enter your task here.*
- **Success:** *Please enter the metric to measure the success of your model here.*

# Data Understanding
Load data and show its structure. Minimum requirements:

- Analysis of missing values.
- 2 visualizations with takeaways.

```{r}
# Load data
data <- data("penguins", package = "palmerpenguins")

# Shows only first 6 rows of the data frame
head(penguins)

# Number of rows and columns
dim(penguins)  # 344 rows and 8 columns

# Check the data formats
str(penguins)

# Check for duplicated values
duplicated_rows <- duplicated(penguins)
# Print the result
if (any(duplicated_rows)) {
  print("There are duplicated rows.")
} else {
  print("No duplicated rows found.")
}

# Summary statistics
summary(penguins)

#interactive spreadsheet-like display of the data frame in RStudio
# View(penguins)

# Calculate the count of missing values per column
missing_count <- colSums(is.na(penguins))
# Calculate the share of missing values per column
missing_share <- colMeans(is.na(penguins))
missing_summary <- data.frame(
        count = missing_count,
        share = round(missing_share, 4)
)
print(missing_summary)

# Class balance for our target population (Gentoo)
gentoo_overview <- penguins %>%
        filter(species == "Gentoo")
cat("Gentoo penguins:", nrow(gentoo_overview), "\n")
table(gentoo_overview$sex, useNA = "ifany")
```

# Explorative Data Analysis
Minimum requirements:

- Distribution of numeric variables
- Relationship between 2 numeric variables
```{r}

# Filter to Gentoo, drop NAs in measurement & sex columns for plotting only
gentoo_eda <- penguins %>%
        filter(species == "Gentoo") %>%
        drop_na(bill_length_mm, bill_depth_mm,
                flipper_length_mm, body_mass_g, sex)

# --- Plot 1: Distribution of all 4 numeric measurements by sex ---
gentoo_long <- gentoo_eda %>%
        select(sex, bill_length_mm, bill_depth_mm,
               flipper_length_mm, body_mass_g) %>%
        pivot_longer(cols = -sex,
                     names_to = "measurement",
                     values_to = "value")

ggplot(gentoo_long, aes(x = sex, y = value, fill = sex)) +
        geom_boxplot(alpha = 0.7) +
        facet_wrap(~ measurement, scales = "free_y") +
        labs(title = "Distribution of physical measurements by sex (Gentoo penguins)",
             subtitle = "Which features separate male from female most clearly?",
             x = "Sex",
             y = "Measurement value") +
        theme_minimal() +
        theme(legend.position = "none")

# --- Plot 2: Relationship between two key features, colored by sex ---
ggplot(gentoo_eda, aes(x = body_mass_g, y = bill_length_mm, color = sex)) +
        geom_point(size = 2.5, alpha = 0.7) +
        labs(title = "Body mass vs. bill length in Gentoo penguins, by sex",
             subtitle = "A 2D view of the strongest separating features",
             x = "Body mass (g)",
             y = "Bill length (mm)",
             color = "Sex") +
        theme_minimal()

```

**Takeaways (bullets):**

- Pattern 1: *Gentoo penguins show clear sexual dimorphism: males are noticeably heavier (median ≈ 5500g vs. 4700g for females) and have longer, deeper bills. The cleanest separation appears in body_mass_g and bill_depth_mm, with bill_length_mm showing modest separation and flipper_length_mm overlapping more.*
- Pattern 2: *The 2D scatterplot of body mass vs. bill length shows two largely separable clusters of male and female penguins, with only a narrow overlap zone — suggesting a simple decision-tree model should perform well using just a few features.*
- Missing values/outliers: *Within Gentoo, 5 of 124 rows have missing values (mostly in sex); these are dropped during preparation. No extreme outliers are visible in the boxplots.*

# Data Preparation
Minimum requirements:

- Drop or impute missing values (explain why)
- Convert categorical variables to factors
```{r}

```
*Note your choices in 1–2 sentences (what you dropped/kept and why).*

# Modeling (pick classification or clustering and delete the other section)

## Classification
Minimum requirements:

- Train/test split (80/20)
- One classification model (e.g., logistic regression, decision tree)
- Report KPI to evaluate
```{r}

```

*1–2 sentences: which classes confuse and why?*

# Evaluation & Validation
Minimum requirements:

- **Metric:** *Please report final metric on test set.*
- **Interpretation (2–4 sentences):** *Please enter the strengths, failure modes (e.g., which species are confused), class imbalance etc here.*

# AI Assistance (brief)
- **Tools used:** *Please enter the tools you used (e.g., GitHub Copilot, ChatGPT) here
- Claude Opus 4.7 Adaptive.*
- **One prompt:** *Please paste one representative prompt that led to useful code here.*
- **How validated:** *Please describe how you checked correctness (e.g., sanity checks on metrics/plots) of the prompt outputs here.*