---
title: "CRISP-DM Skills Check (Penguins)"
author: "Alona Lagodzia, Karoline Kiliyilathu, Julio Dinis"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(42)

# selects the mirror.
chooseCRANmirror(ind=81)

#Run once and then comment out:
# install.packages(c("pandoc"))
# library(pandoc)
# pandoc_install("3.7.0.2")
# chooseCRANmirror(ind=81)

#Run once and then comment out:
# install.packages(c("palmerpenguins","tidyverse","ggplot2", "caret", "rpart", "rpart.plot", "randomForest", "e1071", "class", "kknn", "gbm", "xgboost", "nnet", "pROC", "MLmetrics"))

# load libraries
library(palmerpenguins)
library(tidyverse)
library(ggplot2)

# TODO: use only libraries needed for this work and remove unused ones.
# Add any other libraries you need here:
library(caret)          # train/test split, confusionMatrix,unified model interface
library(rpart)          # Decision Tree
library(rpart.plot)     # nice tree visualization
library(randomForest)   # Random Forest
library(e1071)          # Naïve Bayes + SVM
library(class)          # KNN (basic version)
library(kknn)           # KNN via caret
library(gbm)            # Gradient Boosting Trees
library(xgboost)        # alternative gradient boosting (faster, more popular)
library(nnet)           # simple Neural Networks (built-in style)

# Evaluation extras
library(pROC)           # ROC curves (slide 11 of your lecture)
library(MLmetrics)      # F1 score helper if confusionMatrix isn't enough

```

# Executive summary
**This assignment examines whether the sex of Gentoo penguins can be predicted using physical measurements such as bill length, bill depth, flipper length, and body mass. After cleaning the dataset and removing missing values, a decision tree model was trained using an 80/20 train-test split. The model achieved an accuracy of 0.913 and an F1 score of 0.923, significantly outperforming a baseline of approximately 52%, with body mass and bill depth identified as the most important features. These results indicate that physical characteristics contain strong predictive information for distinguishing between male and female penguins. This suggests that even simple, interpretable models can effectively capture underlying biological differences.**

# Framing the Problem
- **Question:** *To what extent can physical measurements predict the sex of Gentoo penguins, and which features are most informative?*
- **Data:** Palmer Penguins dataset (3 species, 7 variables, 344 rows).
- **Modeling task:** *Binary classification — predicting `sex` (female / male) from four physical measurements (`bill_length_mm`, `bill_depth_mm`, `flipper_length_mm`, `body_mass_g`) using a decision tree (`rpart`). The tree's split structure also reveals which features are most informative, directly addressing the second part of the research question.*
- **Success:** *Accuracy as the primary metric, since the two classes are roughly balanced within Gentoo penguins. F1-score is reported alongside as a robustness check. We compare results to a majority-class baseline (~50% accuracy) to confirm that the model captures real signal beyond chance.*

# Data Understanding
```{r}
# Load data
data <- data("penguins", package = "palmerpenguins")

# Shows only first 6 rows of the data frame
head(penguins)

# Number of rows and columns
dim(penguins)  # 344 rows and 8 columns

# Check the data formats
str(penguins)

# Check for duplicated values
duplicated_rows <- duplicated(penguins)
# Print the result
if (any(duplicated_rows)) {
  print("There are duplicated rows.")
} else {
  print("No duplicated rows found.")
}

# Summary statistics
summary(penguins)

#interactive spreadsheet-like display of the data frame in RStudio
# View(penguins)

# Calculate the count of missing values per column
missing_count <- colSums(is.na(penguins))
# Calculate the share of missing values per column
missing_share <- colMeans(is.na(penguins))
missing_summary <- data.frame(
        count = missing_count,
        share = round(missing_share, 4)
)
print(missing_summary)

# Class balance for our target population (Gentoo)
gentoo_overview <- penguins %>%
        filter(species == "Gentoo")
cat("Gentoo penguins:", nrow(gentoo_overview), "\n")
table(gentoo_overview$sex, useNA = "ifany")
```

# Explorative Data Analysis

```{r}

# Filter to Gentoo, drop NAs in measurement & sex columns for plotting only
gentoo_eda <- penguins %>%
        filter(species == "Gentoo") %>%
        drop_na(bill_length_mm, bill_depth_mm,
                flipper_length_mm, body_mass_g, sex)

# --- Plot 1: Distribution of all 4 numeric measurements by sex ---
gentoo_long <- gentoo_eda %>%
        select(sex, bill_length_mm, bill_depth_mm,
               flipper_length_mm, body_mass_g) %>%
        pivot_longer(cols = -sex,
                     names_to = "measurement",
                     values_to = "value")

ggplot(gentoo_long, aes(x = sex, y = value, fill = sex)) +
        geom_boxplot(alpha = 0.7) +
        facet_wrap(~ measurement, scales = "free_y") +
        labs(title = "Distribution of physical measurements by sex (Gentoo penguins)",
             subtitle = "Which features separate male from female most clearly?",
             x = "Sex",
             y = "Measurement value") +
        theme_minimal() +
        theme(legend.position = "none")

# --- Plot 2: Relationship between two key features, colored by sex ---
ggplot(gentoo_eda, aes(x = body_mass_g, y = bill_length_mm, color = sex)) +
        geom_point(size = 2.5, alpha = 0.7) +
        labs(title = "Body mass vs. bill length in Gentoo penguins, by sex",
             subtitle = "A 2D view of the strongest separating features",
             x = "Body mass (g)",
             y = "Bill length (mm)",
             color = "Sex") +
        theme_minimal()

```

**Takeaways (bullets):**

- Pattern 1: *Gentoo penguins show clear sexual dimorphism: males are noticeably heavier (median ≈ 5500g vs. 4700g for females) and have longer, deeper bills. The cleanest separation appears in body_mass_g and bill_depth_mm, with bill_length_mm showing modest separation and flipper_length_mm overlapping more.*
- Pattern 2: *The 2D scatterplot of body mass vs. bill length shows two largely separable clusters of male and female penguins, with only a narrow overlap zone — suggesting a simple decision-tree model should perform well using just a few features.*
- Missing values/outliers: *Within Gentoo, 5 of 124 rows have missing values (mostly in sex); these are dropped during preparation. No extreme outliers are visible in the boxplots.*

# Data Preparation

```{r}
# Filter to Gentoo, drop incomplete rows, ensure factors
gentoo_clean <- penguins %>%
        filter(species == "Gentoo") %>%
        drop_na() %>%
        mutate(
                sex     = factor(sex),
                species = factor(species),
                island  = factor(island)
        )

# Verification of the cleaned dataset
dim(gentoo_clean)
table(gentoo_clean$sex)
summary(gentoo_clean)
```
*We restrict the dataset to Gentoo penguins (the population of our research question) and drop rows with any missing values, leaving 119 complete observations. Categorical variables (sex, species, island) are converted to factors so they can be used directly by the decision-tree model. Imputation is not used: missingness is small (~4% of Gentoos), and simple deletion is more transparent and avoids introducing assumptions.*

# Modeling

## Classification

```{r}
# Select target and predictor variables
model_data <- gentoo_clean %>%
        select(sex, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g)

# Stratified 80/20 train-test split
set.seed(42)
split_index <- createDataPartition(model_data$sex, p = 0.8, list = FALSE)

train_data <- model_data[split_index, ]
test_data  <- model_data[-split_index, ]

# Verify split and class balance
dim(train_data)
dim(test_data)

table(train_data$sex)
table(test_data$sex)

# Majority-class baseline
baseline_class <- names(which.max(table(train_data$sex)))
baseline_accuracy <- mean(test_data$sex == baseline_class)

cat("Majority-class baseline accuracy:", round(baseline_accuracy, 3), "\n")

# Fit decision tree model
tree_model <- rpart(
        sex ~ bill_length_mm + bill_depth_mm + flipper_length_mm + body_mass_g,
        data = train_data,
        method = "class"
)

# Visualize the decision tree
rpart.plot(
        tree_model,
        type = 4,
        extra = 104,
        main = "Decision tree: predicting Gentoo penguin sex"
)

# Predict on test set
test_preds <- predict(tree_model, newdata = test_data, type = "class")

# Confusion matrix
cm <- confusionMatrix(test_preds, test_data$sex, positive = "male")
cm

# Extract metrics
accuracy <- cm$overall["Accuracy"]
f1_score <- cm$byClass["F1"]

cat("Test accuracy:", round(accuracy, 3), "\n")
cat("F1 score:", round(f1_score, 3), "\n")

# Feature importance
tree_model$variable.importance

importance_df <- data.frame(
        variable = names(tree_model$variable.importance),
        importance = as.numeric(tree_model$variable.importance)
)

ggplot(importance_df, aes(x = reorder(variable, importance), y = importance)) +
        geom_col() +
        coord_flip() +
        labs(
                title = "Feature importance from decision tree",
                x = "Predictor",
                y = "Importance score"
        ) +
        theme_minimal()

```

*The model is a decision tree trained on 80% of the cleaned Gentoo dataset and evaluated on the remaining 20%. Accuracy is used as the primary metric because the classes are approximately balanced, while F1-score is included as a robustness check.*


# Evaluation & Validation

- **Metric:** *The model achieves an accuracy of 0.913 and an F1 score of 0.923 on the test set, clearly outperforming the baseline accuracy of approximately 0.52.*

- **Interpretation:** *The model performs well on a balanced test set (11 female, 12 male), correctly classifying all male penguins and misclassifying only 2 female penguins. Feature importance shows that body mass (33.35) and bill depth (25.32) are the most influential predictors, consistent with patterns observed in the EDA. These results indicate that the model captures the main structure of the data. A limitation is the small test set (23 observations), so results should be interpreted as an estimate rather than a definitive measure.



# AI Assistance
- **Tools used:** *Claude Opus 4.7 Adaptive; GitHub Copilot (GPT-5.2-Codex); Chat GPT*
- **One prompt:** *We are using the Palmer Penguins dataset for a Fundamentals of Data Science homework. Our question is whether physical measurements can predict the sex of Gentoo penguins. Help us implement a simple decision tree model in R with an 80/20 train-test split, confusion matrix, accuracy, F1 score, and feature importance.*
- **How validated:** *We validated the AI-assisted code and text by running the full R Markdown file without errors, checking that the 80/20 train-test split preserved class balance, comparing the model accuracy to a majority-class baseline, and inspecting the confusion matrix. We also checked that the feature importance results were consistent with the EDA plots, especially the visible separation by body mass and bill depth. In addition, AI tools were used for rephrasing, grammar correction, and improving the clarity and consistency of the written text. All AI-assisted content was manually reviewed to ensure it accurately reflects the analysis results and aligns with the assignment requirements and lecture material.*
